---
title: "Getting Started with DOYPAColors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with DOYPAColors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(DOYPAColors)
```

## Overview

The `DOYPAColors` package offers a variety of color palettes designed to enhance your data visualizations in R. These palettes are compatible with `ggplot2` and are categorized into three main types: **sequential**, **diverging**, and **qualitative**. The package also includes a section of colorblind-friendly palettes. This vignette provides a quick guide on how to use the package, including how to list available palettes, preview them, and apply them to your plots.

## Installation

To get started with `DOYPAColors`, you can install the package from CRAN:

```r
install.packages("DOYPAColors")
library(DOYPAColors)
```

Or you can install the development version from GitHub:

```r
if (!require("devtools")) 
  install.packages("devtools")

devtools::install_github("jmestret/DOYPAColors")
library(DOYPAColors)
```

## How to Use

`DOYPAColors` is designed to simplify the process of selecting color palettes for your plots. You can quickly apply a palette without overthinking the details by using the `doypa()` function to automatically choose a palette that fits the type you need, and then use `scale_fill_doypa()` or `scale_color_doypa()` in your `ggplot2` plots.

### Basic Usage with Automatic Palette Selection

Simply specify the type of palette you want (`seq`, `div`, `qual`) and let `DOYPAColors` handle the rest. For example:

```{r}
library(ggplot2)

# Plot using automatic palette selection
ggplot(mtcars, aes(x = factor(cyl), fill = factor(cyl))) +
  geom_bar() +
  scale_fill_doypa(type = "qual", discrete = TRUE) +
  theme_minimal() +
  labs(title = "Bar Plot with Automatic DOYPAColors Palette")
```

In this example, `scale_fill_doypa(type = "qual")` automatically selects a qualitative palette suitable for the number of categories in your data.

For additional options and customization, refer to the rest of this vignette.

## Exploring Palettes

### Listing Available Palettes

To view all available palettes, use the `list_doypa_pals()` function:

```{r}
# List all available DOYPAColors palettes
list_doypa_pals()
```

### Previewing Palettes

You can preview all available palettes using the `preview_doypa_pals()` function:

```{r}
# Preview all available DOYPAColors palettes
preview_doypa_pals()
```

To preview a single palette with multiple plots, use the `preview_pal()` function:

```{r}
# Preview a specific DOYPAColors palette
preview_pal(palette = "doypa")
```

### Getting Palette Colors

To retrieve a vector of colors from a specific palette, use the `doypa()` function. You can specify the palette you want by name, or simply call the function without arguments to get a default palette. If you're unsure which palette to use, let `DOYPAColors` choose one for you:

```{r}
# Retrieve colors from a specific palette
doypa_colors <- doypa(palette = "retro")
print(doypa_colors)

# Retrieve a default palette if no palette name is provided
default_palette_colors <- doypa()
print(default_palette_colors)
```

In the first example, replace `"retro"` with the name of the palette you want to use. If you omit the `palette` argument, `doypa()` will return the colors of a default palette.

## Applying Palettes to ggplot2

### Basic Usage

Integrate DOYPAColors palettes into your `ggplot2` plots using `scale_fill_doypa()` for fill aesthetics and `scale_color_doypa()` for color aesthetics. Here’s an example of how to apply a qualitative palette (`type = "qual"`) to a bar plot:

```{r}
library(ggplot2)

# Create a bar plot with a DOYPAColors palette
ggplot(mtcars, aes(x = factor(cyl), fill = factor(cyl))) +
  geom_bar() +
  scale_fill_doypa(type = "qual", discrete = TRUE) +
  theme_minimal() +
  labs(title = "Bar Plot with DOYPAColors")
```

### Using Palette Options

#### Number of Colors (`n`)

Specify the number of colors to use from the palette with the `n` argument:

```{r}
# Apply a palette with a specific number of colors
ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot() +
  scale_fill_doypa(palette = "buzz", n = 3, discrete = TRUE) +
  theme_classic()
```

#### Reversing the Palette (`reverse`)

Reverse the color order with the `reverse` argument:

```{r, fig.show="hold", out.width="50%"}
ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot() +
  scale_fill_doypa(palette = "buzz", reverse = TRUE, discrete = TRUE) +
  theme_classic() +
  ggtitle("Palette Reversed")

ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot() +
  scale_fill_doypa(palette = "buzz", n = 3, reverse = TRUE, discrete = TRUE) +
  theme_classic() +
  ggtitle("Palette Reversed (n = 3)")
```

#### Creating a Color Gradient (`gradient`)

Generate a color gradient that interpolates between the colors of the selected palette:

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot() +
  scale_fill_doypa(palette = "buzz", gradient = TRUE, discrete = TRUE) +
  theme_classic()
```

## Conclusion

The `DOYPAColors` package makes it easy to apply diverse and visually appealing color palettes to your data visualizations. With options to list, preview, and customize palettes, you have the flexibility to enhance your plots and make them more engaging. Explore the various palettes and options to find the perfect fit for your data.
